---
title: "RNA Seq Part 2"
output: RNA_seq_p2
---

*Run* button within the chunk or by pressing *Cmd+Shift+Enter*. 

*Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

```{r}
library( "DESeq2" )
library(tidyverse)
library(ggplot2)
```

```{r}
# define the path to the file
# note this path is relative to the current working directory and the files are stored in a folder called data
dmso1_data_file_path <- 'rna_counts_data.csv'
# read in the file
dmso_1_count_data <- read.csv(dmso1_data_file_path, header = TRUE)
# show the first few lines
head(dmso_1_count_data)
```
#renames the second column to DMS0_1
```{r}
colnames(dmso_1_count_data)[2] ="DMSO_1"
```

#loads new csv and renames the second column to DMS0_2
```{r}
dmso_2_count_data <- read.csv('rna_counts_data.csv', header = TRUE)
colnames(dmso_2_count_data)[2] ="DMSO_2"
```

#loads new csv and renames the second column to TCMOH_1
```{r}
tcpmoh_1_count_data <- read.csv('rna_counts_data.csv', header = TRUE)
colnames(tcpmoh_1_count_data)[2] ="TCPMOH_1"
```

#loads new csv and renames the second column to TCMOH_2
```{r}
tcpmoh_2_count_data <- read.csv('rna_counts_data.csv', header = TRUE)
colnames(tcpmoh_2_count_data)[2] ="TCPMOH_2"
```

#creates a list of the data frames from all four csvs
```{r}
df_list <- list(dmso_1_count_data, dmso_2_count_data, tcpmoh_1_count_data, tcpmoh_2_count_data)
```

#Merges all the data in teh list into a data frame
```{r}
count_data <- df_list %>% reduce(full_join, by='Geneid')
```

#removes row names and replaces with geneid
```{r}
count_data <- count_data %>% remove_rownames %>% column_to_rownames(var="Geneid")
head(count_data)
```

#creates a table for the metadata of the samples linking sample to its condition
```{r}
# get the sample names from the count_data matrix
SampleName <- c(colnames(count_data))
# specify the conditions for each sample
# In my sample names, DMSO_1 and DMSO_2 are control replicated and TCPMOH_1 and TCPMOH_2 are treated replicates
condition <- c("control", "control", "treated", "treated")
# generate the metadata data frame
meta_data <- data.frame(SampleName, condition)
# make the sample name the row id
meta_data <- meta_data %>% remove_rownames %>% column_to_rownames(var="SampleName")
meta_data
```

#ensures that sample names both in count_data and meta_data 
```{r}
all(colnames(count_data) %in% rownames(meta_data))
```

#ensures that sample names are in the same order in both count_data and meta_data 
```{r}
all(colnames(count_data) == rownames(meta_data))
```

```{r}
# create deseq data set object
dds <- DESeqDataSetFromMatrix(countData = count_data,
                              colData = meta_data,
                              design = ~ condition)
#outputs data set
dds
```

#filters then outputs new counts
```{r}
# filter any counts less than 10
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds
```

```{r}
# set the reference to be the control
dds$condition <- relevel(dds$condition, ref = 'control')
```

```{r}
# get normalized counts
dds <- estimateSizeFactors(dds)
normalized_counts <- counts(dds, normalized = TRUE)

# If you would like to write the normalized counts to a file, you can run the following command. Note that you need to specify the file path. 
# write.table(normalized_counts, file=normalized_counts_data_file_path, sep = '\t', quote=F, col.names = NA)
```

#runs the differential expression pipline
```{r}
dds <- DESeq(dds)
res <- results(dds)
res
```

#summarizes the statistics of the results
```{r}
summary(res)
```

#counts the number of significantly differentially expressed genes below 0.1
```{r}
sum(res$padj < 0.1, na.rm=TRUE)
```

#counts the number of significantly differentially expressed genes below 0.05
```{r}
sum(res$padj < 0.05, na.rm=TRUE)
```

```{r}
# convert results data to basic dataframe
data <- data.frame(res)
head(data)
```

#applies variance stabilizing transformation and creates PCA
```{r}
rld <- vst(dds)
plotPCA(rld)
```

```{r}
# add an additional column that identifies a gene as unregulated, downregulated, or unchanged
# note the choice of pvalue and log2FoldChange cutoff. 
data <- data %>%
  mutate(
  Expression = case_when(log2FoldChange >= log(1) & padj <= 0.05 ~ "Up-regulated",
  log2FoldChange <= -log(1) & padj <= 0.05 ~ "Down-regulated",
  TRUE ~ "Unchanged")
  )
head(data)
```

```{r}
top <- 10
# we are getting the top 10 up and down regulated genes by filtering the column Up-regulated and Down-regulated and sorting by the adjusted p-value. 
top_genes <- bind_rows(
  data %>%
  filter(Expression == 'Up-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top),
  data %>%
  filter(Expression == 'Down-regulated') %>%
  arrange(padj, desc(abs(log2FoldChange))) %>%
  head(top)
  )
# create a datframe just holding the top 10 genes
Top_Hits = head(arrange(data,pvalue),10)
Top_Hits
```

# basic volcano plot showing log2 change vs -log10 change
```{r}
data$label = if_else(rownames(data) %in% rownames(Top_Hits), rownames(data), "")
p1 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point( size = 2/5) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
xlim(-4.5, 4.5)
p1
```

```{r}
# basic plot with line + red for p < 0.05
p2 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point(aes(color = Expression), size = 2/5) +
#geom_hline(yintercept= -log10(0.05), linetype="dashed", linewidth = .4) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
scale_color_manual(values = c("firebrick3", "black", "firebrick3")) +
xlim(-4.5, 4.5) +
theme(legend.position = "none")
p2
```

```{r}
# with labels for top 10 sig overall
library(ggrepel)
p3 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point(aes(color = Expression), size = 2/5) +
# geom_hline(yintercept=-log10(0.05), linetype="dashed", linewidth = .4) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
scale_color_manual(values = c("firebrick3", "black", "firebrick3")) +
xlim(-4.5, 4.5) +
theme(legend.position = "none") +
geom_text_repel(aes(label = label), size = 2.5)
p3
```

```{r}
# plot with up/down
p4 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point(aes(color = Expression), size = 2/5) +
#geom_hline(yintercept=-log10(0.05), linetype="dashed", linewidth = .4) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
scale_color_manual(values = c("dodgerblue3", "black", "firebrick3")) +
xlim(-4.5, 4.5)
p4
```

#final version of volcano plot with gene labels and finalized styling
```{r}
p5 <- ggplot(data, aes(log2FoldChange, -log(pvalue,10))) + # -log10 conversion
geom_point(aes(color = Expression), size = 2/5) +
# geom_hline(yintercept=-log10(0.05), linetype="dashed", linewidth = .4) +
xlab(expression("log"[2]*"FC")) +
ylab(expression("-log"[10]*"P-Value")) +
scale_color_manual(values = c("dodgerblue3", "black", "firebrick3")) +
xlim(-4.5, 4.5) +
geom_text_repel(aes(label = label), size = 2.5)
p5
```

